Supplement to Example Manuscript Template for a Data Analysis Project

### 2024-08-06 code/analysis-code/statistical_analysis.qmd
add scatterplots and update Manuscript

## Import Libraries & Data
```{python}
import pandas as pd
from sklearn.model_selection import train_test_split
import re
pd.set_option('display.max_rows', None)

@@

@@

rand_for_feature_importance_list = rand_for_feature_importance['Feature'].head(10).tolist()
print(rand_for_feature_importance_list)

>>>>

## Import Libraries & Data
```{python}
import pandas as pd
from sklearn.model_selection import train_test_split
from sklearn.preprocessing import StandardScaler
import re
pd.set_option('display.max_rows', None)

@@

scalar = StandardScaler()
scale = scalar.fit_transform(X_grouped)
print(scale)

@@

rand_for_feature_importance_list = xgboost_feature_importance['Feature'].head(10).tolist()
print(rand_for_feature_importance_list)

### 2024-08-06 code/eda-code/eda-ev-tax.qmd
add scatterplots and update Manuscript

# Splitting the data into training and testing sets
X_train_grouped, X_test_grouped, y_train_grouped, y_test_grouped = train_test_split(
    X_grouped, y_grouped, test_size=0.3, random_state=42)

@@

# Preprocess and log transform the training and test datasets
X_train_transformed = preprocess_and_log_transform_data(X_train_grouped)
X_test_transformed = preprocess_and_log_transform_data(X_test_grouped)

@@

    for feature in data.columns:
        fig, ax = plt.subplots(1, 2, figsize=(10,10))
        ax1 = plt.subplot(2,1,1)
        ax1 = sns.histplot(data[feature], kde=True, bins=50)
        ax2 = plt.subplot(2,1,2)
        ax2 = sns.histplot(data1[feature], kde=True, bins=50)
        ax1.title.set_text(f'Histogram of {feature}')
        ax2.title.set_text(f'Log-Transformed Histogram of {feature}')
        plt.xlabel(feature)
        plt.ylabel('Frequency')
        plt.grid(True)
        plt.savefig(f'../../results/figures/histograms/{feature}_histogram.png')

# Plot the histograms for the log-transformed features
plot_transformed_feature_histograms(X_train_grouped, X_train_transformed)
```

@@

    for feature in data.columns:
        fig, ax = plt.subplots(1, 2, figsize=(10,10))
        ax1 = plt.subplot(2,1,1)
        ax1 = sns.boxplot(data[feature])
        ax2 = plt.subplot(2,1,2)
        ax2 = sns.boxplot(data1[feature])
        ax1.title.set_text(f'Boxplot of {feature}')
        ax2.title.set_text(f'Log-Transformed Boxplot of {feature}')
        plt.xlabel(feature)
        plt.ylabel('Frequency')
        plt.grid(True)
        plt.tight_layout()
        plt.savefig(f'../../results/figures/boxplots/{feature}_boxplot.png')

# Plot the boxplots for the log-transformed features
plot_transformed_feature_boxplots(X_train_grouped, X_train_transformed)
```

>>>>

# Splitting the data into training and testing sets
X_train_grouped, X_test_grouped, y_train_grouped, y_test_grouped = train_test_split(
    X_grouped, y_grouped, test_size=0.2, random_state=42)

merged_data.to_csv('test.csv', index=False)
```

```{python}
import matplotlib.pyplot as plt

# Scatter plots to understand relationship between predictors and independent variable
for i, feature in enumerate(features_to_use):
    plt.figure(figsize=(10,5))
    plt.scatter(X_grouped[feature], y_grouped, c="lightblue", edgecolors="blue")
    plt.title(f'{feature} vs Vehicle Count')
    plt.xlabel(feature)
    plt.ylabel('Vehicle Count')
    # plt.grid(True)
    plt.savefig(f'../../results/figures/scatterplots/{feature}_scatterplot.png')

@@

# Preprocess and log transform the training and test datasets
X_transformed = preprocess_and_log_transform_data(X_grouped)

@@

    for feature in data.columns:
        fig, ax = plt.subplots(1, 2, figsize=(10,5))
        ax1 = plt.subplot(2,1,1)
        ax1 = sns.histplot(data[feature], kde=True, bins=50)
        ax2 = plt.subplot(2,1,2)
        ax2 = sns.histplot(data1[feature], kde=True, bins=50)
        ax1.title.set_text(f'Original vs. Log-Transformed: {feature}')
        # ax2.title.set_text(f'Log-Transformed Histogram of {feature}')
        plt.xlabel(feature)
        plt.ylabel('Frequency')
        plt.grid(True)
        plt.savefig(f'../../results/figures/histograms/{feature}_histogram.png')

# Plot the histograms for the log-transformed features
plot_transformed_feature_histograms(X_grouped, X_transformed)
```

@@

    for feature in data.columns:
        fig, ax = plt.subplots(1, 2, figsize=(10,5))
        ax1 = plt.subplot(2,1,1)
        ax1 = sns.boxplot(data[feature])
        ax2 = plt.subplot(2,1,2)
        ax2 = sns.boxplot(data1[feature])
        ax1.title.set_text(f'Original vs. Log-Transformed Boxplot of {feature}')
        # ax2.title.set_text(f'Log-Transformed Boxplot of {feature}')
        plt.xlabel(feature)
        plt.ylabel('Frequency')
        plt.grid(True)
        plt.tight_layout()
        plt.savefig(f'../../results/figures/boxplots/{feature}_boxplot.png')

# Plot the boxplots for the log-transformed features
plot_transformed_feature_boxplots(X_grouped, X_transformed)
```

### 2024-08-06 products/manuscript/Manuscript.qmd
add scatterplots and update Manuscript

The results highlight the Random Forest model's effectiveness in handling diverse and complex datasets, making it a valuable tool for predicting vehicle counts in different zip codes. This study provides critical insights into the practical applications of machine learning in urban planning and resource management, particularly in the context of transportation and infrastructure development.

@@

#### Data Preparation
The dataset was compiled by merging electric vehicle registration data with IRS tax data, focusing specifically on zip codes within the state of Washington. This geographical focus provided a diverse socio-economic landscape, enhancing the robustness of our analysis. The dataset included features such as 'Net premium tax credit amount,' 'Home mortgage interest paid amount,' 'Real estate taxes amount,' and several others related to income, tax credits, and deductions.

@@

#### Feature Transformation
Given that many of the predictor variables exhibited skewed distributions, we applied a log transformation to normalize these features. This transformation was particularly necessary for financial features, such as 'Salaries and wages amount' and 'Home mortgage interest paid amount,' which spanned several orders of magnitude. The transformation formula used was log⁡(x+1)log(x+1), which effectively handled zero values and stabilized variance across the dataset. Normalizing the data in this manner helped improve the performance and interpretability of our machine learning models by reducing the impact of outliers and facilitating a more uniform distribution of feature values.

#### Model Development
We implemented four different machine learning models: Decision Tree, Random Forest, XGBoost, and HistGradientBoosting. Each model was selected for its unique strengths in handling structured data and complex relationships between features.

#### Model Training and Validation
Each model was trained on the processed dataset, with training and validation sets created using cross-validation techniques. Cross-validation helped ensure that the models generalize well to unseen data, preventing overfitting. We evaluated model performance using metrics such as Root Mean Squared Error (RMSE) and R-squared, which provided insights into the accuracy and explanatory power of each model.

@@

### Model Performance Comparison
#### XGBoost
  - RMSE: 71.27 <br/>
  - R-squared: 0.91 <br/>
  - XGBoost demonstrated strong predictive capabilities, achieving a relatively low RMSE of 71.27. The model's R-squared value of 0.91 indicates that it explains 91% of the variance in the vehicle count data, suggesting good model fit and predictive accuracy. <br/>
![XGBoost](../../results/output/model_results/XGBoost_Regressor.png)

#### Decision Tree
  - RMSE: 100.77<br/>
  - R-squared: 0.83<br/>
  -  The Decision Tree model had the highest RMSE among the evaluated models at 100.77, reflecting less precise predictions. Its R-squared value of 0.83 indicates that it explains 83% of the variance, making it the least accurate model in this comparison.
![Decision Tree](../../results/output/model_results/Decision_Tree_Regressor.png)

#### HistGradientBoosting
  -  RMSE: 83.42<br/>
  -  R-squared: 0.88<br/>
  -  The HistGradientBoosting model offered a moderate performance with an RMSE of 83.42 and an R-squared value of 0.88. While better than the Decision Tree model, it still lagged behind XGBoost and Random Forest in terms of prediction accuracy and variance explanation.
![HistGradientBoosting](../../results/output/model_results/HistGradientBoosting_Regressor.png)

#### Random Forest
  -  RMSE: 62.16<br/>
  -  R-squared: 0.935<br/>
  -  The Random Forest model outperformed all other models, achieving the lowest RMSE of 62.16 and the highest R-squared value of 0.935. These metrics indicate that the Random Forest model not only provided the most accurate predictions but also explained 93.5% of the variance in vehicle counts, making it the most reliable model in this analysis.
![Random Forest](../../results/output/model_results/Random_Forest_Regressor.png)

@@

Based on the comparative analysis of the models, the Random Forest model emerged as the best performing model. It achieved the lowest RMSE and the highest R-squared, demonstrating superior accuracy and generalization capability compared to the other models. The Random Forest model's ensemble nature allows it to effectively capture complex patterns in the data, reduce overfitting through bootstrapping, and handle interactions between predictor variables more adeptly. These attributes collectively contributed to its superior performance, making it the most suitable choice for predicting vehicle counts by zip code in this study.

The excellent performance of the Random Forest model suggests it is well-suited for applications requiring high prediction accuracy and robustness, particularly in urban planning and resource allocation scenarios where accurate vehicle count predictions are crucial.

## Discussion
The objective of this study was to develop machine learning models capable of predicting vehicle counts by zip code, using various socio-economic and demographic features derived from IRS tax data. The analysis identified the Random Forest model as the most accurate and robust among the evaluated models, including XGBoost, Decision Tree, and HistGradientBoosting. In addition to the model performance, a detailed examination of the top features contributing to the model predictions provides valuable insights into the socio-economic factors influencing vehicle counts, particularly for electric vehicles (EVs).

### Key Feature Analysis

The top 10 features contributing to the model predictions include:

    Net Premium Tax Credit Amount: This feature represents the tax credit amount received by individuals under the Affordable Care Act. A higher tax credit may indicate lower income levels, suggesting a potential relationship between income and the ability to purchase vehicles, including EVs.

    Home Mortgage Interest Paid Amount: This feature reflects the interest paid on home mortgages, often correlating with homeownership and wealth. Areas with higher mortgage interest payments may have a higher concentration of middle to upper-income households, which can influence vehicle ownership patterns, including the adoption of EVs.

    Real Estate Taxes Amount: Real estate taxes are indicative of property values and, indirectly, wealth. Higher real estate taxes often correspond to higher property values, which may correlate with a higher prevalence of EV ownership due to greater financial capacity.

    Self-Employed Health Insurance Deduction Amount: This deduction is available to self-employed individuals who pay their health insurance premiums. It reflects the presence of self-employed individuals in a region, who may have different financial and vehicle ownership patterns compared to salaried employees.

    Salaries and Wages Amount: This feature directly indicates income levels from employment. Higher aggregate salaries and wages in a zip code suggest greater disposable income, which can facilitate the purchase of vehicles, including newer and potentially more expensive EVs.

    Limited State and Local Taxes: This feature represents the capped amount of state and local taxes that can be deducted from federal taxes. The deduction cap, especially in high-tax states, can impact disposable income and financial planning, potentially influencing vehicle purchasing decisions.

    Home Mortgage from Personal Seller Amount: This feature indicates transactions involving home mortgages from personal sellers. While less common, these transactions can indicate unique financial arrangements in certain areas, which might correlate with specific economic profiles affecting vehicle ownership.

    Individual Retirement Arrangement (IRA) Payments Amount: Contributions to IRAs reflect financial planning and the presence of disposable income. Areas with higher IRA payments may have wealthier populations, potentially increasing the likelihood of EV adoption due to greater financial resources.

    Self-Employed (Keogh) Retirement Plans Amount: Similar to IRAs, Keogh plans are retirement savings plans for self-employed individuals. Contributions to these plans suggest higher income and financial planning, which can correlate with higher rates of vehicle ownership, including EVs.

    Additional Medicare Tax Amount: This tax applies to high-income earners and indicates a population segment with substantial earnings. Higher additional Medicare tax payments correlate with higher income levels, potentially leading to increased EV adoption due to greater financial capacity.

### Implications and Socio-Economic Insights

The analysis of these top features reveals a consistent pattern: higher vehicle counts, particularly for EVs, are associated with higher income levels, wealth indicators (such as mortgage interest and real estate taxes), and financial planning (such as retirement contributions). These factors suggest that regions with greater financial capacity are more likely to adopt EVs, reflecting broader trends in consumer behavior and market accessibility.

The emphasis on features related to self-employment and additional taxes for high earners highlights the impact of income diversity on vehicle ownership patterns. The presence of deductions and credits for self-employed individuals suggests that entrepreneurial regions may exhibit unique vehicle ownership trends, possibly favoring flexibility and lower-cost vehicles or, conversely, reflecting higher disposable income that supports EV adoption.

@@

{{< pagebreak >}}
## Citations
Jain, A. (n.d.). Electric vehicle population [Data set]. Kaggle. Retrieved June 6, 2024, from https://www.kaggle.com/datasets/jainaru/electric-vehicle-population <br/>

Internal Revenue Service. (n.d.). SOI tax stats - Individual income tax statistics - ZIP code data (SOI). Retrieved June 16, 2024, from https://www.irs.gov/statistics/soi-tax-stats-individual-income-tax-statistics-zip-code-data-soi

>>>>

The results highlight the XGBoost model's effectiveness in handling diverse and complex datasets, making it a valuable tool for predicting vehicle counts in different zip codes. This study provides critical insights into the practical applications of machine learning in urban planning and resource management, particularly in the context of transportation and infrastructure development.

@@

### Data Preparation & Exploratory Analysis
The dataset was compiled by merging electric vehicle registration data with IRS tax data, focusing specifically on zip codes within the state of Washington. This geographical focus provided a diverse socio-economic landscape, enhancing the robustness of our analysis. The dataset included features such as 'Net premium tax credit amount,' 'Home mortgage interest paid amount,' 'Real estate taxes amount,' and several others related to income, tax credits, and deductions.

@@

#### Scatterplots - Predictor's vs. Vehicle Count Relationship
![SE Health Insurance Deduction Scatterplot](../../results/figures/scatterplots/Self-employed%20health%20insurance%20deduction%20amount_scatterplot.png)

### Feature Transformation
Given that many of the predictor variables exhibited skewed distributions, we applied a log transformation to normalize these features. This transformation was particularly necessary for financial features, such as 'Salaries and wages amount' and 'Home mortgage interest paid amount,' which spanned several orders of magnitude. The transformation formula used was log⁡(x+1)log(x+1), which effectively handled zero values and stabilized variance across the dataset. Normalizing the data in this manner helped improve the performance and interpretability of our machine learning models by reducing the impact of outliers and facilitating a more uniform distribution of feature values.

#### Log-Transform Histogram
![SE Health Insurance Deduction](../../results/figures/histograms/Self-employed%20health%20insurance%20deduction%20amount_histogram.png)

#### Log-Transform Boxplot
![SE Health Insurance Deduction Boxplot](../../results/figures/boxplots/Self-employed%20health%20insurance%20deduction%20amount_boxplot.png)

### Model Development
We implemented four different machine learning models: Decision Tree, Random Forest, XGBoost, and HistGradientBoosting. Each model was selected for its unique strengths in handling structured data and complex relationships between features.

### Model Training and Validation
Each model was trained on the processed dataset, with training and validation sets created using cross-validation techniques. Cross-validation helped ensure that the models generalize well to unseen data, preventing overfitting. We evaluated model performance using metrics such as Root Mean Squared Error (RMSE) and R-squared, which provided insights into the accuracy and explanatory power of each model.

@@

### Model Performance Comparison
#### XGBoost
  - RMSE: 69.21 <br/>
  - R-squared: 0.892 <br/>
  - XGBoost demonstrated strong predictive capabilities, achieving a relatively low RMSE of 71.27. The model's R-squared value of 0.91 indicates that it explains 89.2% of the variance in the vehicle count data, suggesting good model fit and predictive accuracy. <br/>
![XGBoost](../../results/output/model_results/XGBoost_Regressor.png)

#### Decision Tree
  - RMSE: 79.61<br/>
  - R-squared: 0.857<br/>
  -  The Decision Tree model had the highest RMSE among the evaluated models at 79.61, reflecting less precise predictions. Its R-squared value of 0.857 indicates that it explains 85.7% of the variance, making it the least accurate model in this comparison.
![Decision Tree](../../results/output/model_results/Decision_Tree_Regressor.png)

#### HistGradientBoosting
  -  RMSE: 76.96<br/>
  -  R-squared: 0.867<br/>
  -  The HistGradientBoosting model offered a moderate performance with an RMSE of 76.96 and an R-squared value of 0.867. While better than the Decision Tree model, it still lagged behind XGBoost and Random Forest in terms of prediction accuracy and variance explanation.
![HistGradientBoosting](../../results/output/model_results/HistGradientBoosting_Regressor.png)

#### Random Forest
  -  RMSE: 71.30<br/>
  -  R-squared: 0.885<br/>
  -  The Random Forest model outperformed all other models, achieving the lowest RMSE of 71.30 and the highest R-squared value of 0.885. These metrics indicate that the Random Forest model not only provided the most accurate predictions but also explained 88.5% of the variance in vehicle counts, making it the most reliable model in this analysis.
![Random Forest](../../results/output/model_results/Random_Forest_Regressor.png)

@@

The XGBoost model emerged as the best performer in predicting vehicle counts by zip code, primarily due to its advanced boosting techniques and the ability to handle various data complexities effectively. The combination of these features makes XGBoost a powerful and versatile tool for predictive modeling. Its ability to handle complex, structured data, prevent overfitting, and efficiently process large datasets, while offering detailed feature importance analysis, made it the top performer in this study. The model's superior performance in terms of low RMSE and high R-squared values underscores its effectiveness in accurately predicting vehicle counts by zip code, providing valuable insights for urban planning and policy-making.

## Discussion
The top 10 features identified as the most significant predictors of vehicle counts by zip code offer insights into the socio-economic and financial characteristics of the population. These features provide a nuanced understanding of the factors that potentially influence vehicle ownership, particularly the adoption of electric vehicles (EVs). Here's a detailed discussion of each feature and its relevance:

## Variable Importance - XGBoost
!['XGBoost'](../../results/output/variable_importance/XGBoost.png)

1. Self-Employed Health Insurance Deduction Amount
This feature represents the amount self-employed individuals can deduct for health insurance premiums. It is significant as it indicates the presence and financial activities of self-employed individuals within a zip code. Self-employed individuals often have variable incomes and unique financial situations, which may influence their ability to purchase and maintain vehicles. This deduction could correlate with areas where self-employment is prevalent, possibly reflecting regions with more entrepreneurial activity and, potentially, higher disposable incomes.

2. Individual Retirement Arrangement Payments Amount
Contributions to Individual Retirement Arrangements (IRAs) reflect financial planning and savings behavior. Higher contributions suggest higher disposable income and financial stability, factors that are positively associated with the ability to purchase vehicles, including more expensive options like EVs. This feature indicates regions where residents prioritize long-term financial health, which may correlate with the capacity to invest in newer, environmentally friendly technologies.

3. Home Mortgage from Personal Seller Amount
This feature involves transactions where the home mortgage is obtained directly from a personal seller rather than a traditional financial institution. It indicates non-standard financial arrangements, possibly reflecting unique housing markets or economic conditions in certain areas. The relevance of this feature could be tied to the accessibility and flexibility in financial transactions within these regions, potentially influencing overall economic activity, including vehicle purchases.

4. Salaries and Wages Amount
A direct indicator of income, this feature represents the total salaries and wages reported within a zip code. It is a critical predictor as higher salaries and wages generally correlate with greater financial capacity to purchase and maintain vehicles. This feature strongly influences vehicle ownership levels, including the adoption of EVs, which may have higher upfront costs but offer long-term savings.

5. Home Mortgage Interest Paid Amount
The amount of interest paid on home mortgages is indicative of both property ownership and the financial burden of maintaining such properties. Higher mortgage interest payments suggest higher home values and, by extension, potentially higher wealth levels. This wealth is a significant factor in vehicle ownership, as individuals with more expensive homes often have higher disposable incomes, enabling them to invest in additional vehicles or more luxurious options, including EVs.

6. Self-Employment Tax Amount
This feature captures the amount of self-employment tax paid, which directly correlates with self-employment income. Higher self-employment tax payments indicate successful businesses or high-earning self-employed individuals. The economic health of these individuals likely contributes to their ability to afford vehicles, making this a significant predictor of vehicle counts.

7. Self-Employed (Keogh) Retirement Plans Amount
Contributions to Keogh plans reflect savings set aside by self-employed individuals for retirement. Like IRA contributions, this feature suggests a focus on long-term financial security, indicative of higher income levels and financial planning. This financial stability is associated with greater capacity for discretionary spending, including vehicle purchases.

8. Business or Professional Net Income (Less Loss) Amount
This feature represents the net income from business or professional activities, after accounting for losses. It is a measure of the economic success and profitability of businesses within a zip code. Higher net incomes suggest prosperous business environments, which can lead to increased vehicle ownership among business owners and employees, reflecting economic prosperity in these areas.

9. Real Estate Taxes Amount
Real estate taxes are directly linked to property values and indirectly to wealth. Higher real estate taxes generally indicate more expensive properties, suggesting higher overall wealth in an area. Wealthier populations have greater access to credit and financing options, which can increase the likelihood of purchasing vehicles, including newer and more expensive models like EVs.

10. Net Premium Tax Credit Amount
The net premium tax credit amount provides a tax credit under the Affordable Care Act, generally aimed at reducing the cost of health insurance for lower to middle-income families. While it may seem counterintuitive, this feature could correlate with regions where financial assistance is necessary, reflecting economic diversity. In these areas, the adoption of vehicles, including EVs, may be influenced by financial incentives or subsidies aimed at promoting energy-efficient technologies.

@@

{{< pagebreak >}}
## Citations

### 2024-08-06 products/manuscript/supplement/Supplementary-Material.qmd
add scatterplots and update Manuscript

# Additional results
## Log-Transformed Histogram Examples
![SE Health Insurance Deduction](../../../results/figures/histograms/Self-employed%20health%20insurance%20deduction%20amount_histogram.png)

@@

![Net Premium Tax Credit Amount](../../../results/figures/histograms/Net%20premium%20tax%20credit%20amount_histogram.png)

## Log-Transformed Boxplot Examples
![SE Health Insurance Deduction](../../../results/figures/boxplots/Self-employed%20health%20insurance%20deduction%20amount_boxplot.png)

@@

![Net Premium Tax Credit Amount](../../../results/figures/boxplots/Net%20premium%20tax%20credit%20amount_boxplot.png)

@@

>>>>

# Additional results
## Log-Transformed Histogram Examples
![Individual Retirement Payments](../../../results/figures/scatterplots/Individual%20retirement%20arrangement%20payments%20amount_scatterplot.png)

@@

![Home Mortgage Interest Paid](../../../results/figures/histograms/home%20mortgage%20interest%20paid%20amount_histogram.png)

## Log-Transformed Boxplot Examples
![Individual Retirement Payments](../../../results/figures/boxplots/Individual%20retirement%20arrangement%20payments%20amount_boxplot.png)

@@

![Home Mortgage Interest Paid](../../../results/figures/boxplots/home%20mortgage%20interest%20paid%20amount_boxplot.png)

@@

## Variable Importance - Decision Tree
!['Decision Tree Regressor'](../../../results/output/variable_importance/Decision%20Tree%20Regressor.png)

## Variable Importance - Hist Gradient Boosting
!['Hist Gradient Boosting'](../../../results/output/variable_importance/HistGradientBoostingRegressor%20(Permutation%20Importance).png)

## Variable Importance - Random Forest
!['Random Forest'](../../../results/output/variable_importance/Random%20Forest%20Regressor.png)
